---
title: "Adjunct Instruction at SLCC"
author: Jeff Webb
date: \today

output:
  ioslides_presentation:
    incremental: true
    widescreen: true
    #smaller: true
    css: https://maxcdn.bootstrapcdn.com/bootstrap/3.3.7/css/bootstrap.min.css
---



## Purpose
- Present preliminary findings on adjunct instruction at SLCC.
- Get feedback from domain experts to inform additional work.

## Introduction
- Reliance on part-time faculty at community colleges has been increasing for decades as a percentage of total faculty:
    + 1969: 27%
    + 1989: 52%
    + 2007: 70%
    + 2018: nearly 80% at SLCC
- Why?  Adjuncts are 80% cheaper than full-time faculty! (In 2007 the percentage of adjuncts at 4-year schools was only 32%.)
- **Key question**:  Do adjuncts have a different impact on student outcomes than full-time instructors?


## Definitions

- **Adjunct**: An instructor on a temporary contract who is limited to part-time instruction during the term of the contract.
- **Full-time**:  A permanent instructor who teaches full-time.
- These employment status designations are mutually exclusive within terms and exhaustive:  to my knowledge there are no other possible employment options for instructors at SLCC.

## Key points from the literature {.smaller}

- Compared to full-time faculty, adjuncts:
    + tend to be less engaged in campus culture and less available to students (Jaeger & Eagan 2009)
    + are less likely to have an office on campus or a campus email address (Landrum 2009).
    + have almost half the years of teaching experience (Landrum 2009).
    + earn substantially less (Greenberg 2014).
    + have less academic freedom and feel more hostage to student and administrative opinions (Wilson 2010).
    + have positive impacts on grades earned in introductory courses (Ran 2017)
    + have negative impacts on subsequent course enrollment and performance (Ran 2017)
    + increased the number of subsequent classes students took in a given field, especially for fields closely tied to a profession (Bettinger & Long 2010).


## Research questions

1. **Description**: What was the nature and scope of adjunct teaching at SLCC?
2. **Grades**: Did student grades vary by faculty employment status?
3. **Retention**: Was retention into the next semester different for students who took courses largely from adjunct instructors compared to those who took courses largely from full-time instructors?
4. **Graduation**: Did students who had a high percentage of adjunct instructors, either in the first year or beyond, graduate at lower rates?
5.  **Next course performance**: If students took a beginning course from an adjunct instructor was there a subsequent performance difference in the following course?

## Data notes

- Dataset consisted in one row per student course enrollment from fall 2011 through spring 2018, for a total of 20 academic terms.  
- Vocational enrollments were excluded as well as concurrent enrollments.  
- Only students in VFA Fall cohorts were included in the data.  


# Descriptive Results


## Data discrepancy


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=F, warning=F,cache=TRUE)

# packages and data

library(tidyverse)
library(knitr)
library(kableExtra)
library(ggthemes)


d <- read.csv("TASK0019261 Adjuct Instructor Research-2.txt",
              header=T, sep="|", stringsAsFactors = F)

names(d) <- tolower(names(d))

d$class <- paste0(d$course_subject, d$course_number)

d$unique_classID <- paste0(d$course_subject, d$course_number,
                           d$course_section, d$term_code)

# add total enrollment in a particular class to the main data
d <- d %>%
  group_by(class) %>%
  summarize(total_enrollment = n()) %>%
  left_join(y=d, by="class")


# add total enrollment in a particular section to the main data
d <- d %>%
  group_by(unique_classID) %>%
  summarize(total_section_enrollment = n()) %>%
  left_join(y=d, by="unique_classID")

# Create a new data frame with by filtering and adding features
subd <- d %>%
  filter(school != "SAT" &
           #total_enrollment>3000 &
           academic_vocational=="Academic Course" &
           full_time_ind != "U" &
           instruction_type != "Lab" &
           instruction_type != "NOT USED - Lab" &
           substr(term_code, 5,6) != "50" &
           final_grade != "MC" &
           final_grade != "NM" &
           final_grade != "NS" &
           final_grade != "MG" &
           final_grade != "P" &
           final_grade != "") %>%
  mutate(grad_term = substr(slcc_graduation_term, 1, 6),
         inst_status = ifelse(full_time_ind == "Y", "Full-time", "Adjunct"),
         online = ifelse(campus=="SLCC Online", "online", "in-person"),
         course_level = ifelse(course_number < 1000, "Developmental",
                               ifelse(course_number >= 1000 & course_number < 2000,
                                      "1000 level", "2000 level")),
         course_level = factor(course_level,
                               levels = c("Developmental", "1000 level", "2000 level")),
         term = ifelse(substr(term_code, 5,6)=="30", "Summer",
                       ifelse(substr(term_code, 5,6)=="40", "Fall", "Spring")),
         term = factor(term, levels = c("Summer", "Fall", "Spring")),
         prior_ug_gpa_ch = as.character(cut(prior_ug_gpa, breaks = quantile(prior_ug_gpa, probs = seq(0,1, .1), na.rm=T),
                                                   include.lowest = T)),
         prior_ug_credits_ch = as.character(cut(prior_ug_credits, breaks = quantile(prior_ug_credits, probs = seq(0,1, .1), na.rm=T),
                                                   include.lowest = T))) 


subd$prior_ug_gpa_ch[is.na(subd$prior_ug_gpa_ch)] <- "none"
subd$prior_ug_credits_ch[is.na(subd$prior_ug_credits_ch)] <- "none"

# any(is.na(subd$prior_ug_gpa_ch))
# any(is.na(subd$prior_ug_credits_ch))

# letter grade numeric equivalents
grades <- data.frame(final_grade = c("A", "A-","B+","B","B-","C+","C",
                                     "C-","D+","D","D-", "E","W","I"),
                     alt_grade= c(4,3.7, 3.4,3,2.7,2.4,2,1.7,1.4,1,.7,0,0,0))

# add numeric equivalents to subd data frame
subd <- subd %>%
  left_join(grades, by= "final_grade")

# create pass variable set at 3
subd$pass <- ifelse(subd$alt_grade >= 3, 1, 0)

# Create groups
# table(d$term_code, d$academic_year)

terms <- data.frame(terms = unique(subd$term_code)) %>% arrange(terms)

instructors <- subd %>%
  filter(inst_status == "Adjunct") %>%
  group_by(instructor_id) %>%
  tally %>%
  dplyr::select(instructor_id)

instructor_df <- expand.grid(term_code = terms$terms, instructor_id = instructors$instructor_id)

groups <- subd %>%
  filter(inst_status=="Adjunct") %>%
  group_by(instructor_id, term_code) %>%
  tally %>%
    dplyr::select(-n) %>%
  arrange(instructor_id, term_code) %>%
  group_by(instructor_id) %>%
   mutate(n = seq_along(instructor_id)) %>%
  right_join(instructor_df, by = c("instructor_id", "term_code")) %>%
  group_by(instructor_id) %>%
  mutate(adjunct_group = ifelse(all(is.na(head(n, 2))) & all(is.na(tail(n, 2))) & max(n, na.rm=T)==1, "Short term",
                                ifelse(all(is.na(head(n, 2))) & all(is.na(tail(n, 2))) & max(n, na.rm=T) > 1 & max(n, na.rm=T) < 6, "Medium term",
                                       ifelse(max(n, na.rm=T) > 5, "Long term", NA)))) %>%
  na.omit %>%
  filter(adjunct_group == "Short term" |
           adjunct_group=="Medium term" & n > 1 |
         adjunct_group =="Long term" & n > 5,
         !(term_code %in% c(201140, 201220,  201820, 201740))) %>%
  rename(adjunct_group_n = n)

subd <- subd %>%
  left_join(groups, by = c("instructor_id", "term_code"))




# Plotting functions

theme_slcc <- function(base_size=12) {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size)
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2)),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(),
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               legend.key.size= unit(0.2, "cm"),
               legend.margin = margin(0, unit="cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))

}

scale_fill_slcc <- function(...){
      library(scales)
      discrete_scale("fill","slcc",manual_pal(values = c("#00ABE1", "#FFCD00", "#003865", "#833921")), ...)

}

scale_colour_slcc <- function(...){
      library(scales)
      discrete_scale("colour","slcc",manual_pal(values = c("#00ABE1", "#FFCD00", "#003865", "#833921")), ...)

}

colors=c("#00ABE1", "#FFCD00", "#003865", "#833921")


ipeds <- read.csv("ipeds_ft_pt_counts.csv")



# summary of adjunct teaching in 17-18
perc_faculty <- d %>%
  filter(academic_year=="2017-2018",
         academic_vocational=="Academic Course") %>%
  group_by(instructor_id, full_time_ind) %>%
  tally %>%
  #filter(full_time_ind != "U") %>%
  group_by(full_time_ind) %>%
  summarize(count= n()) %>%
  mutate(perc= round(count/sum(count),2))

percent_students_taught <- d %>%
  filter(academic_year=="2017-2018",
         academic_vocational=="Academic Course") %>%
  group_by(instructor_id, full_time_ind) %>%
  tally %>%
  group_by(full_time_ind) %>%
  summarize(count= sum(n)) %>%
  mutate(perc= round(count/sum(count),2))

# perc_faculty
# percent_students_taught

```



```{r}
data.frame(Data = c("IPEDS","SLCC IR","IPEDS","SLCC IR","IPEDS","SLCC IR"),
             Term = c("Fall 2014", "Fall 2014","Fall 2015", "Fall 2015","Fall 2016","Fall 2016"),
           Fulltime = c(343, 287, 339,300, 335, 294),
           Parttime = c(1155, 1064, 1271, 1049, 1242, 1040)) %>%
  dplyr::select(`Data source`=Data, Term, `Full-time`= Fulltime, `Part-time`=Parttime) %>%
  mutate(Total = `Full-time` + `Part-time`,
         `% part-time` = round(`Part-time`/Total*100,1)) %>%
  kable(caption = "Full-time vs. Part-time SLCC faculty by data source, Fall terms 2014 - 2016", escape = F) %>%
  kable_styling("striped", full_width = T)
```

## IPEDS data

```{r}

# fig:ipeds
ipeds %>%
  mutate(`2016` = pt_2016/(pt_2016 + ft_2016)*100,
         `2015` = pt_2015/(pt_2015 + ft_2015)*100,
          `2014` = pt_2014/(pt_2014 + ft_2014)*100) %>%
  dplyr::select(instnm, `2016`,`2015`,`2014`) %>%
  gather(year, pt_proportion, -instnm) %>%
  group_by(year) %>%
  mutate(slcc = pt_proportion[instnm=="Salt Lake Community College"]) %>%
  ggplot(aes(pt_proportion)) +
  geom_histogram(bins=10, fill = colors[1])+
  geom_vline(aes(xintercept = slcc), col=2, lty = 2) +
  facet_wrap(~year) +
  theme_minimal() +
  labs(x = "Percentage of part-time faculty",
       title = "Percentage of part-time faculty at 38 large community colleges",
       subtitle = "SLCC percentage indicated by the red dashed line",
       caption = "Data source:  IPEDS") +
  theme_slcc()


```

<!-- ## Adjuncts as percentage of faculty by academic year (IR data) -->
<!-- ```{r} -->

<!-- subd %>% -->
<!--   group_by(academic_year, instructor_id, inst_status) %>% -->
<!--  tally %>% -->
<!--   filter(!(academic_year %in% c("2011-2012", "2017-2018"))) %>% -->
<!--   group_by(academic_year, inst_status) %>% -->
<!--   summarize(n = n()) %>% -->
<!--   group_by(academic_year) %>% -->
<!--   mutate(percent = round(n/sum(n)*100,1)) %>% -->
<!--   filter(inst_status == "Adjunct") %>% -->
<!--   select(`Academic Year` = academic_year, Percent = percent) %>% -->
<!--   kable() %>% -->
<!--   kable_styling("striped")  -->

<!-- ``` -->

## SLCC instructor demographics

```{r}
# Instructor table
# fixed characteristics

inst_age_table <- subd %>%
  distinct(instructor_id, inst_status,instructor_age) %>%
  group_by(instructor_id, inst_status) %>%
  summarize(age = mean(instructor_age)) %>%
  mutate(cat_age = ifelse(age >=10 & age<20, "[10-20)",
                          ifelse(age >= 20 & age< 30, "[20-30)",
                                 ifelse(age >=30 & age<40, "[30-40)",
                                        ifelse(age >=40 &age <50, "[40-50)",
                                               ifelse(age >=50 & age< 60, "[50-60)",
                                                      ifelse(age >=60 & age < 70, "[60-70)", ">70")))))),
         cat_age = factor(cat_age, levels=c("[10-20)","[20-30)", "[30-40)", "[40-50)", "[50-60)", "[60-70)", ">70"))) %>%
  group_by(cat_age, inst_status) %>%
  tally %>%
  group_by(inst_status) %>%
  mutate(Percent = round(n/sum(n)*100,1))

#inst_age_table

# Gender
inst_gender_table <-subd %>%
  distinct(instructor_id, inst_status, instructor_gender) %>%
  group_by(instructor_gender,inst_status) %>%
  tally %>%
  group_by(inst_status) %>%
  mutate(Percent = round(n/sum(n)*100,1)) #%>%
  #summarize(sum(n))

# inst_gender_table


# Ethnicity
inst_ethnicity_table <-subd %>%
  distinct(instructor_id, inst_status, instructor_ethnicity) %>%
  group_by(instructor_ethnicity, inst_status) %>%
  tally %>%
  group_by(inst_status) %>%
  mutate(Percent = round(n/sum(n)*100,1)) #%>%
  #summarize(sum(n))

# inst_ethnicity_table

# total
inst_total_table <- subd %>%
  distinct(instructor_id, inst_status) %>%
  group_by(inst_status) %>%
  tally %>%
  mutate(sum(n))

# inst_total_table

 
  eth_table <- inst_ethnicity_table %>%
  filter(instructor_ethnicity!="Non-Resident Alien") %>%
   gather(variable, value, -c(instructor_ethnicity, inst_status))%>%
  unite(temp, inst_status, variable) %>%
  spread(temp,value) %>%
  unname %>%
    data.frame 
  
  names(eth_table)  <- c("Ethnicity","Count", "Percent", "Count","Percent")
   
  eth_table %>%
    kable() %>%
  kable_styling("striped", full_width=T) %>%
  add_header_above(c(" ", "Adjunct" = 2, "Full-time" = 2))
```

## SLCC instructor demographics  
  
```{r}

  gender_table <- inst_gender_table %>%
  gather(variable, value, -c(instructor_gender, inst_status))%>%
  unite(temp, inst_status, variable) %>%
  spread(temp,value) %>%
  unname %>%
    data.frame

names(gender_table)  <- c("Gender","Count", "Percent", "Count","Percent")
   
  gender_table %>%
    kable() %>%
  kable_styling("striped", full_width=T) %>%
  add_header_above(c(" ", "Adjunct" = 2, "Full-time" = 2))
```

## SLCC instructor demographics  

```{r}
  age_table <- inst_age_table %>%
  gather(variable, value, -c(cat_age, inst_status))%>%
  unite(temp, inst_status, variable) %>%
  spread(temp,value) %>%
  unname %>%
    data.frame


 names(age_table) <- c("Age","Count", "Percent", "Count","Percent")

 age_table %>%
  kable() %>%
  kable_styling("striped", full_width=T) %>%
 add_header_above(c(" ", "Adjunct" = 2, "Full-time" = 2))
```



## Scope of adjunct teaching at SLCC 

- We can distinguish between the percentage of faculty that are adjunct and the percentage of *student enrollments* taught by adjuncts.   
- Adjunct faculty taught about 64% of student enrollments on average while full-time faculty taught about 36%.  
- Class sizes for adjuncts and full-time instructors were comparable, but adjuncts, being part time, taught fewer classes (on average, 2 per term) than did full-time faculty (4 per term). 
- However, the percentage of student enrollments taught by adjuncts varied by term, with adjuncts teaching about 3% more enrollments in summer.
- The percentage of student enrollments taught by adjuncts varied by department and course.

## Percentage of Student Enrollments Taught by Adjuncts by Department 

```{r}
subd %>%
  group_by(course_subject, inst_status) %>%
  tally %>%
  group_by(course_subject) %>%
  mutate(total = sum(n),
         perc = round(n/sum(n)*100,1)) %>%
  arrange(desc(total)) %>%
  filter(inst_status=="Adjunct")%>%
   head(9) %>%
  dplyr::select(Subject=course_subject, Count=n,Percent = perc) %>%
  kable() %>%
  kable_styling('striped')
```

## Percentage of Student Enrollments Taught by Adjuncts by Course 

```{r}
subd %>%
  group_by(class, inst_status) %>%
  tally %>%
  group_by(class) %>%
  mutate(total = sum(n),
         perc = round(n/sum(n)*100,1)) %>%
  arrange(desc(total)) %>%
  filter(inst_status=="Adjunct")%>%
    head(9) %>%
  dplyr::select(Course=class, Count = n, Percent = perc) %>%
  kable() %>%
  kable_styling('striped')


```

## Additional differences between adjunct and full-time 
- Adjuncts taught  66% of the in-person students but only 51% of the online students.
- Adjuncts also taught almost 70% of the students  enrollments at the 1000-2000 course level, 58% at the developmental level (< 1000), and an even smaller percentage---43%---at the 2000 level.
- The percentage of course enrollments in general education courses taught by adjuncts approached the overall average.  

## Types of Adjuncts

- I categorized adjuncts into groups based on length of service:
    + short-term: those who taught only 1 term and did not teach again.
    + medium-term:  those who taught between 2 and 5 terms and did not teach again.
    + long-term:  those who taught more than 5 terms.

## Summary of adjunct groups

```{r}


subd %>% 
  filter(!is.na(adjunct_group)) %>%
  mutate(n = adjunct_group_n) %>%
  group_by(`Adjunct group` = adjunct_group) %>%
  summarize(`Minimum terms` = min(n),
            `Maximum terms` = max(n),
            `Average terms` = round(mean(n),1),
            Count = length(unique(instructor_id))) %>%
  mutate(Percent = round(Count/sum(Count)*100, 1),
         `Adjunct group` = factor(`Adjunct group`, levels = c("Short term", "Medium term", "Long term"))) %>%
  kable() %>%
  kable_styling('striped')

```

## Grades


- Adjunct instructors awarded higher grades, on average, compared to full-time faculty. 
- However, the raw average difference, unadjusted for other factors, was modest: 2.53 for adjuncts compared to 2.39 for full-time faculty. 
- The most noticeable difference occurred at 0 and 4.
- Note:  0 includes I, W and E.

## Grades

```{r }

subd %>%
  group_by(inst_status, alt_grade) %>%
  summarize(count = n()) %>%
  group_by(inst_status) %>%
  mutate(total= sum(count),
         perc = count/total*100) %>%
  dplyr::select(`Instructor status` =inst_status, alt_grade, count, total, perc) %>%
  ggplot(aes(alt_grade, perc, fill= `Instructor status`)) +
  geom_bar(stat="identity", position = "dodge") +
  labs(x = "Final grades",
       y = "Percentage",
       title = "Distribution of grades by instructor employment status",
       subtitle = "Academic classes only, 2011 - 2018") +
  theme_slcc() +
  scale_fill_slcc()

```

## Pass rate

- Passing:  numeric grade greater than or equal to 3. 
- Adjuncts had higher pass rates:  56.4% vs. 51.6%. 
- Summer pass rates were higher than those for Fall or Spring---almost 60% compared to 54%---due to the fact that adjuncts taught relatively more in the Summer.
- The overall 5 point difference might seem modest, but it was large in absolute terms.
    + If full-time instructors had used adjunct standards for grading then over 7500 students from this dataset would have passed who did not.
    + If adjuncts had used the grading standards of full-time instructors then over 13,000 students would not have passed who did.
    
# Statistical Results

## Grades

- Did student grades differ by instructor employment status?
- There were many sources of grade variation---such as the difficulty of the course or the  delivery method---that needed to be disentangled from the variation associated with instructor employment status.
- **Result**: The average difference in grades uniquely associated with instructor employment status was .07 [.02, .12], with adjuncts awarding slightly higher grades.
- However, this difference varied significantly by course.
- Courses with a large difference between full-time and adjunct grading might benefit from faculty-wide grade norming or developing shared course materials.

## Grades 

```{r include = F}

library(lme4)
library(arm)

mm3 <- lmer(alt_grade ~ online +
             term+
       gen_ed +
       instructor_gender +
       instructor_ethnicity +
       instructor_age +
       student_gender +
       student_ethnicity +
       student_age_by_term + 
       prior_ug_credits_ch +
       prior_ug_gpa_ch + 
       inst_status + 
         (1 + inst_status | class), 
       data = subset(subd, total_enrollment > 3000))

se <- se.ranef(mm3)
re <- ranef(mm3)
class_var <- data.frame(class = row.names(data.frame(re$class)),
                          est = data.frame(re$class)[,2],
                        lower = data.frame(re$class)[,2] - 2*data.frame(se$class)[,2],
                          upper = data.frame(re$class)[,2] + 2*data.frame(se$class)[,2])


int <- tail(fixef(mm3),1) 
```


```{r}

ggplot(class_var, aes(reorder(class, est), est + int)) +
  geom_point() +
  geom_errorbar(aes(ymin= lower + int, ymax = upper + int) )+
  theme_slcc() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Courses",
       title = "Grade differences by instructor employment status and course (> 3000)",
       subtitle = "Point estimates and 95% confidence intervals for full-time > adjunct grades",
       y = "Average grade difference") +
  geom_hline(lty=2, col=2, yintercept = 0)


```


## Retention

- Did student next term retention vary by instructor employment status?  
- This question was not straightforward to answer because students typically took multiple courses in any given term taught by a mix of full-time and adjunct instructors. 
- So, to rephrase:  Did the probability of being retained the following semester change as the proportion of courses taken from adjuncts increased?  
- This question was difficult to answer because retention itself is pretty random.  This makes it hard to be certain about small sources of variation in retention.

## Retention

```{r include = F}


(prev_retention <- subd %>%
    filter(vfa_cohort_term > 201130) %>%
  group_by(student_id, term_code) %>%
  slice(1) %>%
    group_by(student_id) %>%
    mutate(term_seq = seq_along(student_id),
           cum_retain = cumsum(ifelse(returned_next_term_ind=="Y",1,0)),
           prop_retain = cum_retain/term_seq,
           prev_retain = c("first term", round(prop_retain[-length(student_id)],1))) %>%
    dplyr::select(student_id, term_code, term_seq, prev_retain))


  
  # data

returned <- subd %>%
  filter(vfa_cohort_term > 201130, 
         term == "Fall") %>%
  left_join(prev_retention, by = c("student_id", "term_code")) %>%
   arrange(student_id, term_code) %>%
  group_by(student_id, term_code) %>%
  summarize(course_n = n(),
            first_term = ifelse(all(term_code==vfa_cohort_term),1,0),
            term = first(term),
            adjunct = sum(ifelse(full_time_ind=="N", 1, 0)),
            proportion_adjunct = adjunct/course_n,
            avg_course_number = mean(course_number),
            max_course_number = max(course_number),
            prior_ug_credits = first(prior_ug_credits_ch),
            math = ifelse(any(course_subject=="MATH"),1,0),
            english = ifelse(any(course_subject=="ENGL"),1,0),
            proportion_gened = mean(ifelse(gen_ed=="Y",1,0)),
            age = first(student_age_by_term),
            ethnicity = first(student_ethnicity),
            gender = first(student_gender),
            retained = first(ifelse(returned_next_term_ind=="Y", 1, 0)), 
            prior_gpa = first(prior_ug_gpa_ch),
            term_gpa = mean(alt_grade, na.rm=T),
            online = mean(ifelse(online=="online",1,0)),
            term_seq = first(term_seq),
            prev_retain = paste0("[",first(prev_retain),"]"))

#in first term
r4 <- glm(retained ~ course_n +first_term*proportion_adjunct +  
            term_gpa, 
          family = binomial, 
          data = returned) 

pred_frame2 <- expand.grid(course_n= 1:6,
                          proportion_adjunct = c(0,1),
                          term_gpa = round(quantile(unique(returned$term_gpa), probs = c(0, .2,.4,.6,.8,1)),2),
                          first_term= c(0,1))



pred_frame2$pred <- predict(r4, newdata= pred_frame2, type = "response")

pred_frame2$gpa_ch <- paste("GPA:", pred_frame2$term_gpa)
pred_frame2$adj_ch <- paste("Proportion adjunct:", pred_frame2$proportion_adjunct)
pred_frame2$first_ch <- paste("First term:", ifelse(pred_frame2$first_term==0, "No","Yes"))

# create probability curves for this result
```


```{r}
pred_frame2 %>%
  filter(proportion_adjunct %in% c(0,1)) %>%
  ggplot(aes(course_n, pred, col = factor(proportion_adjunct))) +
  geom_line() +
  facet_grid(first_ch~gpa_ch) +
  labs(title = "Student retention by term GPA, first term and instructor status",
       subtitle= "Model predicted probability curves for student retention into spring term",
       x = "Number of classes taken",
       y = "Probability of student retention") +
  guides(col=guide_legend(title="Proportion of classes taught  by adjuncts")) +
  theme_slcc() +
  scale_colour_slcc()

``` 


## Retention

```{r}

pd <- read.csv("pd_plot.csv")

ggplot(pd, aes(Feature.Value, Partial.Dependence)) +
  geom_line() + 
  theme_slcc() + 
  labs(x = "Proportion of adjuncts",
       y = "Probability of retention into spring",
       title = "Partial dependence curve from machine learning model of retention")

```


## Retention

- **Result**: The effect of adjunct teaching on retention *appears* to be real, but small, with many countervailing influences.  (I am suspicious, though:  The mechanism for the effect is unknown.)
- Many students do not take classes from only adjuncts or only full-time instructors, and therefore lie somewhere in between the lines in the Figure.  
- Compare two average students in their first term: the first took 1 out 4 courses from an adjunct and the second who took 3 out of 4 from an adjunct.
    + The probability of returning for the next term declines by only about .025 for the second student.
    + If the same comparison is made after the first term the probability declines by only .015.
    
## Graduation

- Did graduation vary by adjunct teaching?  
- Graduation is a career-level outcome, so we need to be more specific: Did the probability of graduating change as the proportion of career courses taken from adjuncts increased, all else being equal?  
- The tricky issue here is how to define "all else being equal."  Adjusting for cumulative credits was important.
- **Result**: There does not appear to be much of a relationship between adjunct instruction and graduation.   

## Next course performance

- Did student performance in a sequence of courses---say, Engl1010 then Engl2010 or MATH1010 then Math1050---vary by the employment status of the first instructor when the second course was taught by a full-time instructor?  
- This question is trying to get at the issue of learning, independently of grading practice.  
- Additionally, what effect might adjunct instruction have had on student retention into that second course?  
- **Result**: Next course performance did not appear to differ much, if at all, by the employment status of the first course instructor.  The estimated effects were small and uncertain.
